\normalfont{A bird's-eye view on the habitability of exoplanets via statistical learning techniques}

### 2020-10-02 Manuscript/Manuscript.tex
PUB

Before applying the algorithm previously described, a preliminary analysis via the factor analysis of mixed data (FAMD) was performed, this allowed to know the variables that were statistically correlated or not. Note that the distances  factors (such as the mean distance $P\_D$, planet apastron $P\_A$ and the planet periastron $P\_PN$) are poorly correlated with the planet habitability: this can be ascribed to the fact that the habitability was extended beyond the "earth zone" by considering habitable planets that, with a particular atmosphere as shown in fig \ref{Planets_habitability_Seager}, have the correct surface temperature for liquid water
\footnote{In principle one may wonder if a better approach should considered only the planets like Earth habitable or perhaps to distinguish the two cases in order to get the distance effect: the issue here is that these exoplanets are very few, thus to preserve the significance of the habitable planets, if such approach was undertaken, a further reduction of the dataset had to be considered. Since this choice strongly limits the quality of forecasts, such approach was not considered.} thus the statistical correlation between the distance factors and the habitability is strongly lowered. The first approach that was considered is the classification decision tree. First the complexity parameter (CP) was checked (a 10-fold cross-validation on the training set \footnote{In this work, for all the approaches undertaken, the training ensemble consist in a set of 350 exoplanets while the test one is composed by 150 items} was undertaken): as can be seen from Fig \ref{Decision_tree_performances} the default parameter $0.01$ can be used. Then the decision tree algorithm was applied to the dataset: as it can be seen from Fig \ref{DecisionTree} the main discriminant is the stellar luminosity: this feature discriminate, as one expects, the largest part of the habitable planet from the non-habitable. The nodes $2$, $4$, $9$, $16$, $33$ and $67$ discriminates deeply the habitability on the basis of stellar temperature ($S_T$), the planet distance from the host star ($P\_D$ mean distance and $P\_PN$ periastron) and the planet temperature equilibrium temperature ($P\_T\_E$). The performance of this algorithm on this dataset, evaluated with a 10-fold cross-validation \footnote{For this procedure the test set and the validation sets where reshuffled 10 times by modifying the numerical seed by which the exoplanets are assigned to the training or to the test set. The proportion 150/500 for test and training sets was keep constant}, is reported in Fig \ref{Decision_tree_performances} and in the table \ref{DecisionTree_perf_tab}. Next the random forest approach was undertaken. As shown in Fig. \ref{Random_forest_tree_perfs} the main parameters for this algorithm were previously checked. Its results are reported in Fig \ref{Random_forest_tree_results}: as one can point out the importance of the variables is slightly changed with respect to the decision tree. Such changes reflects on the performances of the model when the overall CV (as described before) is performed: as one can note from Fig the performances are slightly improved with respect to the ones reported in Fig \ref{Decision_tree_performances}, in particular the ratio of the false positive is reduced with a percentage of 40 $\%$. Such improvement is captured by the accuracy and $\phi$ factors reported in table \ref{Random Forest_perf_tab}. Next the SVM classification algorithm was considered: as the previous cases the most important parameters , in this case the budget and the polynomial degree , were tuned (see Fig. \ref{SVM_tuning}). The performances of this approach are reported in the Fig \ref{SVM_results} and in the table \ref{SVM_perf_tab}: these seems to be worst with respect to the decision tree and random forest. Such behaviour can be ascribed to the large flexibility of the algorithm that produce a slight overfitting. In order to improve the performance of SVM the dimensionality reduction with principal component analysis (PCA) \footnote{Since the planet habitability is a categorical variable a particular library was used: indeed the PCAmixdata package provides the function \textit{PCAmix()} that manage the quantitative variables with the principal component analysis (PCA), while the qualitative variables are undertaken via the multiple correspondence analysis (MCA)} was considered : this allows not only to improve the model, as reported in Fig \ref{SVM+PCA_results} and in the Tab. \ref{SVM+PCA_perf_tab}, but also to plot in to a 2D graph, as in Fig. \ref{PCA+SVM_PLOT} the support vector and the decision boundary.

@@

@@

@@

\begin{figure}[htp]
  \centering
\subfloat[]{\label{Decision_Tree_FINAL_CV}\includegraphics[width=0.6\textwidth]{Pic/Decision_Tree_FINAL_CV.png}}\\
\subfloat[]{\label{Decision_Tree_FINAL_CV}\includegraphics[width=0.6\textwidth]{Pic/Decision_Tree_FINAL_ROC.png}}
\caption{The performances of the decision tree algorithm on the habitability of exoplanets as given by the confusion matrix (upper panel) and ROC curve (lower panel) as obtained by a 10-fold cross validation. An predicted (P) habitable exoplanet is classified as 1 while a test (T) non-habitable exoplanet is classified as 0}
\label{Decision_tree_performances}
\end{figure}

@@

\begin{figure}[h]
  \centering
\subfloat[]{\includegraphics[width=0.6\textwidth]{Pic/Random_Forest_FINAL_CV.png}}\\
\subfloat[]{\includegraphics[width=0.6\textwidth]{Pic/Random_Forest_FINAL_ROC.png}}
\caption{The performances of the random forest algorithm on the habitability of exoplanets as given by the confusion matrix (upper panel) and ROC curve (lower panel) as obtained by a 10-fold cross validation. An predicted (P) habitable exoplanet is classified as 1 while a test (T) non-habitable exoplanet is classified as 0.}
\label{Random_forest_tree_results}
\end{figure}

\begin{figure}[h]
\begin{center}
\includegraphics[width=0.8\textwidth]{Pic/SVM_tuning.png}
\caption{The tuning of the budget parameter and the polynomial degree for the SVM classification algorithm. The selected point represent the parameters chosen for this work}
\label{SVM_tuning}
\end{center}
\end{figure}

\begin{figure}[h]
  \centering
\subfloat[]{\includegraphics[width=0.6\textwidth]{Pic/SVM_FINAL_CV.png}}\\
\subfloat[]{\includegraphics[width=0.6\textwidth]{Pic/SVM_FINAL_ROC.png}}
\caption{The performances of the SVM on the habitability of exoplanets as given by the confusion matrix (upper panel) and ROC curve (lower panel) as obtained by a 10-fold cross validation. An predicted (P) habitable exoplanet is classified as 1 while a test (T) non-habitable exoplanet is classified as 0.}
\label{SVM_results}
\end{figure}

@@

\begin{figure}[h]
\begin{center}
\subfloat[]{\includegraphics[width=0.5\textwidth]{Pic/FAMD_squared_loadings.pdf}}\\
\subfloat[]{\includegraphics[width=0.5\textwidth]{Pic/FAMD_quantitative_variables.pdf}}
\subfloat[]{\includegraphics[width=0.5\textwidth]{Pic/FAMD_Qualitative_VAR.pdf}}\\
\caption{The FAMD analysis of the the dataset: on the panel (a)  the relationship square, on  the panel (b) the correlation circle is provided and on panel (c) the representation of the categories is given.}
\label{PCA+SVM_PLOT}
\end{center}
\end{figure}

\begin{figure}[h]
  \centering
\subfloat[]{\includegraphics[width=0.6\textwidth]{Pic/SVM_PCA_ROC.png}}\\
\subfloat[]{\includegraphics[width=0.6\textwidth]{Pic/SVM_PCA_CV.png}}
\caption{The performances of the SVM on the habitability of exoplanets (as obtained by a 10-fold cross validation) once the the dimensionality reduction via was applied (to 2 dimension): on the panel (a) the confusion matrix is given  while on panel (b) the ROC curve (lower panel) is reported. An predicted (P) habitable exoplanet is classified as 1 while a test (T) non-habitable exoplanet is classified as 0.}
\label{SVM+PCA_results}
\end{figure}

>>>>

Before applying the algorithm previously described, a preliminary survey via the factor analysis of mixed data (FAMD) was performed: this allowed to know the variables that were statistically correlated or not. Note that the distances  factors (such as the mean distance $P\_D$, planet apastron $P\_A$ and the planet periastron $P\_PN$) are poorly correlated with the planet habitability: this can be ascribed to the fact that the habitability was extended beyond the \textit{earth zone} by considering habitable planets that, with a particular atmosphere as shown in fig \ref{Planets_habitability_Seager}, have the correct surface temperature for liquid water
\footnote{In principle one may wonder if a better approach should considered only the planets like Earth habitable or perhaps to distinguish the two cases in order to get the distance effect: the issue here is that these exoplanets are very few, thus to preserve the significance of the habitable planets, if such approach was undertaken, a further reduction of the dataset had to be considered. Since this choice strongly limits the quality of forecasts, such approach was not considered.} thus the statistical correlation between the distance factors and the habitability is strongly lowered. The first approach that was considered is the classification decision tree. First the complexity parameter (CP) was checked (a 10-fold cross-validation on the training set \footnote{In this work, for all the approaches undertaken, the training ensemble consist in a set of 350 exoplanets while the test one is composed by 150 items} was undertaken): as can be seen from Fig \ref{Decision_tree_performances} the default parameter $0.01$ can be used. Then the decision tree algorithm was applied to the dataset: as it can be seen from Fig \ref{DecisionTree} the main discriminant is the stellar luminosity: this feature discriminate, as one expects, the largest part of the habitable planet from the non-habitable. The nodes $2$, $4$, $9$, $16$, $33$ and $67$ discriminate deeply the habitability on the basis of stellar temperature ($S_T$), the planet distance from the host star ($P\_D$ mean distance and $P\_PN$ periastron) and the planet temperature equilibrium temperature ($P\_T\_E$). The performance of this algorithm on this dataset, evaluated with a 10-fold cross-validation \footnote{For this procedure the test set and the validation sets where reshuffled 10 times by modifying the numerical seed by which the exoplanets are assigned to the training or to the test set. The proportion 150/500 for test and training sets was keep constant}, is reported in Fig \ref{Decision_tree_performances} and in the table \ref{DecisionTree_perf_tab}. Next the random forest approach was undertaken. As shown in Fig. \ref{Random_forest_tree_perfs} the main parameters for this algorithm were previously checked. Its results are reported in Fig \ref{Random_forest_tree_results}: as one can point out the importance of the variables is slightly changed with respect to the decision tree. Such change reflects on the performances of the model when the overall CV (as described before) is performed: as one can note from Fig the performances are slightly improved with respect to the ones reported in Fig \ref{Decision_tree_performances}, in particular the ratio of the false positive is reduced with a percentage of 40 $\%$. Such improvement is captured by the accuracy and $\phi$ factors reported in table \ref{Random Forest_perf_tab}. Next the SVM classification algorithm was considered: as the previous cases the most important parameters , in this case the budget and the polynomial degree , were tuned (see Fig. \ref{SVM_tuning}). The performances of this approach are reported in the Fig \ref{SVM_results} and in the table \ref{SVM_perf_tab}: these seems to be worst with respect to the decision tree and random forest. Such behaviour can be ascribed to the large flexibility of the algorithm that produce a slight overfitting. In order to improve the performance of SVM the dimensionality reduction with principal component analysis (PCA) \footnote{Since the planet habitability is a categorical variable a particular library was used: indeed the PCAmixdata package provides the function \textit{PCAmix()} that manage the quantitative variables with the principal component analysis (PCA), while the qualitative variables are undertaken via the multiple correspondence analysis (MCA)} was considered : this allows not only to improve the model, as reported in Fig \ref{SVM+PCA_results} and in the Tab. \ref{SVM+PCA_perf_tab}, but also to plot in to a 2D graph, as in Fig. \ref{PCA+SVM_PLOT} the support vectors and the decision boundary. Finally we considered the liner and quadratic discriminant analysis: their performances are plotted in Fig. and reported in Fig. \ref{LDA+QDA_results} as well in the Tab. \ref{LDA_perf_tab}, \ref{QDA_perf_tab}.

@@

\begin{figure}[h]
\begin{center}
\subfloat[]{\includegraphics[width=0.5\textwidth]{Pic/FAMD_squared_loadings.pdf}}\\
\subfloat[]{\includegraphics[width=0.5\textwidth]{Pic/FAMD_quantitative_variables.pdf}}
\subfloat[]{\includegraphics[width=0.5\textwidth]{Pic/FAMD_Qualitative_VAR.pdf}}\\
\caption{The FAMD analysis of the dataset: on the panel (a)  the relationship square, on  the panel (b) the correlation circle is provided and on panel (c) the representation of the categories is given.}
\label{PCA+SVM_PLOT}
\end{center}
\end{figure}

@@

\begin{table}[]
\caption{Performances of LDA algorithm for the exoplanets habitability, evaluated by the accuracy, the no-information rate (NIR), and the $\phi$ factor.}
\begin{center}
\begin{tabular}{l|l}
\multicolumn{1}{c|}{Feature} & Value \\ \hline
\multicolumn{1}{c|}{Accuracy} & 0.957 \\ \hline
\multicolumn{1}{c|}{NIR} & 0.898 \\ \hline
\multicolumn{1}{c|}{$\phi$} & 0.738 \\ 
\label{LDA_perf_tab}
\end{tabular}
\end{center}
\end{table}

\begin{table}[]
\caption{Performances of QDA algorithm for the exoplanets habitability, evaluated by the accuracy, the no-information rate (NIR), and the $\phi$ factor.}
\begin{center}
\begin{tabular}{l|l}
\multicolumn{1}{c|}{Feature} & Value \\ \hline
\multicolumn{1}{c|}{Accuracy} & 0.957 \\ \hline
\multicolumn{1}{c|}{NIR} & 0.899 \\ \hline
\multicolumn{1}{c|}{$\phi$} & 0.887 \\ 
\label{QDA_perf_tab}
\end{tabular}
\end{center}
\end{table}

@@

\begin{figure}[htp]
  \centering
\subfloat[]{\label{Decision_Tree_FINAL_CV}\includegraphics[width=0.6\textwidth]{Pic/Decision_Tree_FINAL_CV.png}}\\
\subfloat[]{\label{Decision_Tree_FINAL_CV}\includegraphics[width=0.6\textwidth]{Pic/Decision_Tree_FINAL_ROC.png}}
\caption{The performances of the decision tree algorithm on the habitability of exoplanets as given by the confusion matrix (upper panel) and ROC curve (lower panel) as obtained by a 10-fold cross validation. A predicted (P) habitable exoplanet is classified as 1 while a test (T) non-habitable exoplanet is classified as 0}
\label{Decision_tree_performances}
\end{figure}

@@

\begin{figure}[h]
  \centering
\subfloat[]{\includegraphics[width=0.6\textwidth]{Pic/Random_Forest_FINAL_CV.png}}\\
\subfloat[]{\includegraphics[width=0.6\textwidth]{Pic/Random_Forest_FINAL_ROC.png}}
\caption{The performances of the random forest algorithm on the habitability of exoplanets as given by the confusion matrix (upper panel) and ROC curve (lower panel) as obtained by a 10-fold cross validation. A predicted (P) habitable exoplanet is classified as 1 while a test (T) non-habitable exoplanet is classified as 0.}
\label{Random_forest_tree_results}
\end{figure}

\begin{figure}[h]
\begin{center}
\includegraphics[width=0.8\textwidth]{Pic/SVM_tuning.png}
\caption{The tuning of the budget parameter and the polynomial degree for the SVM classification algorithm. The selected point (x represents the budge while x the polynomial degree) is the parameter set chosen for this work}
\label{SVM_tuning}
\end{center}
\end{figure}

\begin{figure}[h]
  \centering
\subfloat[]{\includegraphics[width=0.6\textwidth]{Pic/SVM_FINAL_CV.png}}\\
\subfloat[]{\includegraphics[width=0.6\textwidth]{Pic/SVM_FINAL_ROC.png}}
\caption{The performances of the SVM on the habitability of exoplanets as given by the confusion matrix (upper panel) and ROC curve (lower panel) as obtained by a 10-fold cross validation. A predicted (P) habitable exoplanet is classified as 1 while a test (T) non-habitable exoplanet is classified as 0.}
\label{SVM_results}
\end{figure}

@@

\begin{figure}[h]
  \centering
\subfloat[]{\includegraphics[width=0.6\textwidth]{Pic/SVM_PCA_ROC.png}}\\
\subfloat[]{\includegraphics[width=0.6\textwidth]{Pic/SVM_PCA_CV.png}}
\caption{The performances of the SVM on the habitability of exoplanets (as obtained by a 10-fold cross validation) once the the dimensionality reduction via was applied (to 2 dimension): on the panel (a) the confusion matrix is given  while on panel (b) the ROC curve (lower panel) is reported. A predicted (P) habitable exoplanet is classified as 1 while a test (T) non-habitable exoplanet is classified as 0.}
\label{SVM+PCA_results}
\end{figure}

\begin{figure}[h]
  \centering
\subfloat[]{\includegraphics[width=0.4\textwidth]{Pic/LDA_FULL_CV.png}}
\subfloat[]{\includegraphics[width=0.4\textwidth]{Pic/QDA_FULL_CV.png}}\\
\subfloat[]{\includegraphics[width=0.4\textwidth]{Pic/LDA_FULL_ROC.png}}
\subfloat[]{\includegraphics[width=0.4\textwidth]{Pic/QDA_FULL_ROC.png}}
\caption{The performances of the LDA (left panels a and c) and QDA (right panels b and d) on the habitability of exoplanets as obtained by a 10-fold cross validation.  A predicted (P) habitable exoplanet is classified as 1 while a test (T) non-habitable exoplanet is classified as 0.}
\label{LDA+QDA_results}
\end{figure}
